---
title: "Linear Regression (Frequentist and Bayesian) - 2025-05-06"
author: "Prof. Gwen Eadie"
date: "2025-05-06"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

## Fitting a simple regression line to mock data (frequentist approach)

This exercise follows section 6.2 in Gelman, Hill, and Vehtari (2021).

1.  Simulate 20 fake data points from the model $$y_i = a +bx_i + \epsilon_i$$ where there is one predictor (covariate) that takes on values between 1 and 20, and where the intercept is 0.2 and the slope is 0.3, and $$\epsilon_i \sim N(0, \sigma=0.5)$$

```{r}
# write code and a function to simulate y values here. 
n=20
x = 1:n
simfunc = function(a, b, sigma, x){ a + b*x + rnorm(n, mean = 0, sd = sigma) }

mya = 0.2
myb = 0.3
mysigma = 0.5

y = simfunc(a=mya, b=myb, sigma = mysigma, x=x)
```

2.  Fit a linear regression model to these fake data. You will need to create a `data.frame` containing both the predictor and the outcome. Use the function `lm` to fit the simple linear regression.

```{r}
# create data frame with covariate and outcome in columns
fakedata = data.frame(x, y)

# you should use the lm function and assign the results to a new object
lmfit = lm(y ~ x, data = fakedata)
```

3.  Now display the results using `print`. What does print show you? 

```{r}
print(lmfit)
```
4. Display results using `summary`, which gives a little more information (e.g., the standard error of each estimate)
```{r}
summary(lmfit)

```

5. Plot the data, regression line, and confidence bands
```{r}
# simple plot without confidence bands
plot(fakedata$x, fakedata$y, pch=19, xlab="x", ylab="y")
grid()
abline(lmfit, col="darkgrey")

# can use the visreg package to easily make fit line with confidence bands
library(visreg)
visreg(lmfit)

```


## Inferring the parameters for a line (Bayesian approach)


Using the same mock data, infer the parameters of the intercept and slope.
```{r}
# load packages
library(ggplot2)
library(tidyverse)
library(tidybayes)
library(brms)
```

1.  We will use the default priors in stan_lm.
```{r}

#bayesfit <- stan_lm(y ~ x, data = fakedata, prior=NULL)
# use Bayesian Regression Model (brm) function from brms package
bayesfit <- brm(y ~ x, data = fakedata)

```
2.  Look at the chains and summary of the output of the inferred posterior distribution
```{r}
# plot the Markov chains and marginal distribution of the draws
plot(bayesfit)

# get the summary output
summary(bayesfit)
```

3.  Plot the inferred relationship using the posterior distribution
```{r}
# Create a new data frame that will be used to generate predictions given the (Bayesian) regression prediction of y at a bunch x's
new_data <- data.frame(x = seq(min(fakedata$x), max(fakedata$x), length.out = 100)) 

# Get fitted values for the new_data object
fitted_values <- fitted(bayesfit, newdata = new_data) 

# Use the posterior draws and new_data object to calculate credible intervals
posterior_draws <- posterior_epred(bayesfit, newdata = new_data, draws = 1000)
    
# Calculate the credible interval (e.g., 95% CI):
credible_interval <- apply(posterior_draws, 2, function(x) quantile(x, probs = c(0.025, 0.975))) 


# Now plot the line and credible intervals using the ggplot2 package
ggplot(new_data, aes(x = x)) +
    geom_ribbon(aes(ymin = credible_interval[1,], ymax = credible_interval[2,]), fill = "lightblue", alpha = 0.5) +
    geom_line(aes(y = fitted_values[,1]), color = "blue") +
    geom_point(data = fakedata, aes(y = y)) +
    labs(title = "Regression Line with Credible Bands", x = "x", y = "y") +
    theme_bw()
```

